# DataFrameConst

This **R** package defines two S4 classes 

- `HomogList`: a list in which all elements must be the same class
- `DataFrameConst`: a data frame with optional required columns and 
  classes, or general constraints.

It also defines the most common methods
`[<-`, `[[<-`, `$<-`, `c`, `cbind2`, `rbind2` for these classes so that the constraints
are checked when data in the objects are updated.

## HomogList

Create a list in which all elements must be functions

```{r}
library("DataFrameConstr")
foo <- HomogList(list(sum=sum, max=max, min=min), "function")
print(foo)
```

Since `HomogList` extends `list`, it can be used like any other list.
```{r}
x <- 1:10
lapply(foo, function(f) f(x))
```

It can be updated,
```{r}
foo[["mean"]] <- mean
print(foo)
```
but the object will return an error if an element other than the specified
class name is returned.
```{r}
foo[["a"]] <- 1
```
The methods
`[<-`, `[[<-`, `$<-` and `c` are all defined to return `HomogList` objects, 
and by extension, check the class types of the elements in the new list.

The function `subclass_homog_list` can be used to create subclasses of
`HomogList` for a specified class. The function `subclass_homog_list`, 
will create the class and all its associated methods, and return a
function which creates new objects of that class.

For example, the following creates a new class "FunctionList", in
which all elements must be `function` objects.
```{r}
FunctionList <- subclass_homog_list("FunctionList", "function")
```

Then a new object of class `FunctionList` can be created either by
```{r}
FunctionList(list(sum=sum, mean=mean))
```
or, more verbosely,
```{r}
new("FunctionList", list(sum=sum, mean=mean))
```

What is important about this class is that it will not accept any
non-function elements either on creation,
```{r}
FunctionList(list(a=1))
```
or when updating an existing object,
```{r}
foo <- FunctionList(list(sum=sum, mean=mean))
foo[["a"]] <- 1
```

This makes classes extending `HomogList` particularly useful with S4
objects, either to define lists of S4 objects, or as the slot class
for a class definition.

For example, in the **coda** package, `mcmc.list` is a S3 class
consisting of a list of `mcmc` objects. An equivalent S4 class, which
I'll call `NewMcmcList`, could be created with one function call, 
```r
NewMcmcList <- subclass_homog_list("NewMcmcList", "mcmc.list") 
```

## DataFrameConstr

The `DataFrameConstr` class extends the `data.frame` class, but allows
for required columns with specified classes, and for general
constraints on the `data.frame`. 

For example, let's create a data frame which must have an `numeric` column named
`"a"`, a column named `"b"`, which can be of any class, and a
`"factor` column named `"c"`. Additionally, require that all
values of `"a"` are positive.

```{r}
foo <- 
  DataFrameConstr(data.frame(a = runif(3), b = runif(3), c = letters[1:3]),
                  columns = c(a = "numeric", b = "ANY", c = "factor"),
                  constraints = list(function(x) {x$a > 0}))
```

The new object `foo` acts just like any other `data.frame`,
```{r}
print(foo)
summary(foo)
```

However, it will validate updates to ensure that the data meets the specified constraints,
This will return an error because `a` was defined as `numeric`,
```{r}
foo$a <- as.character(foo$a)
```
This returns an error because `a` is constrained to be strictly positive,
```{r}
foo["a", 1] <- -1
# Unfortunately, this syntax, does not work, and alters foo
# foo[["a"]][1] <- -1
# I can't figure out how to avoid that, so if anyone knows, can you let me know? 
```

This will not cause an error because the column `b` is allowed to have any class (more formally, 
it is of class "ANY"),
```{r}
foo$b <- as.character(foo$b)
```

Since `foo` was created with `exclusive=FALSE` (by default) then the data frame can contain more rows 
than `a`, `b`, and `c`. The following is valid,
```{r}
foo$d <- runif(3)
```
However, `foo` is guaranteed to always contain columns `a`, `b`, and `c`, and thus these columns cannot 
be deleted. This will return an error,
```{r}
foo$a <- NULL
```

The methods
`[<-`, `[[<-`, `$<-`, `cbind2` (use instead of `cbind`), and `rbind2` (use instead of `rbind`), are defined so that they return `DataFrameConstr` objects, and by extension check the column classes, and constraints of the new object.

The function `constrained_data_frame` can be used to create subclasses
of `DataFrameConstr`. For example, to create a class, which I'll call `"Foo"`, which 
has the same columns and constraints as the `foo` object previously created,
```{r}
Foo <- 
  constrained_data_frame("Foo", 
                         columns = c(a = "numeric", b = "ANY", c = "factor"),
                         constraints = list(function(x) {x$a > 0}))
```
Now there is a new class, `"Foo"`, which inherits from `DataFrameConstr`,
```{r}
showClass("Foo")
```

Then create a new object, `bar` of class `Foo`,
```{r}
bar <- Foo(data.frame(a = runif(3), b = runif(3), c = letters[1:3]))
```
This new object will validate any new data, so the following will produce errors,
```{r}
bar[["a"]] <- as.character(bar[["a"]])
bar[["a"]] <- -1
bar[["a"]] <- NULL
```

This will validate the object on creation, so the following will return an error, 
because it does not contain the columns `b` or `c`,
```{r}
Foo(data.frame(a = runif(3)))
```

The additional capabilities that `DataFrameConstr` adds to
`data.frames` make it useful for the following,

- slot class types within S4 objects
- data validation
- creating an **R** ORM to databases





<!--  LocalWords:  runif DataFrameConst HomogList cbind rbind lapply
 -->
<!--  LocalWords:  DataFrameConstr HomogClass homog FunctionList mcmc
 -->
<!--  LocalWords:  NewMcmcList showClass ORM
 -->
